---
title: "HW2: Starbucks Global Store Analysis"
author: "Jihong Min"
date: "2025-02-20"
output: 
  html_document
runtime: shiny
---


```{r, echo = FALSE}
library(knitr)
library(shiny)
library(leaflet)

#opts_knit$set(base_dir = "/", base.url = "/")
opts_chunk$set(
  warning = FALSE,
  message = FALSE
  )
```

```{r code=readLines("https://raw.githubusercontent.com/JIHONGKING/Min/refs/heads/main/starbucks_r.R")}
```

## **Introduction**  

This project explores **Starbucks' global store distribution** using **Shiny**, an interactive web framework for R.  

- **Dataset Source:** [Starbucks Locations Worldwide 2021 (Kaggle)](https://www.kaggle.com/datasets/kukuroo3/starbucks-locations-worldwide-2021-version)  
- **Scope:** Covers **over 28,000 stores across 49 countries**  
- **Key Attributes:** Includes **store locations, ownership types, and operational status**  

## **Objective**  

The goal of this project is to:  
- **Visualize Starbucks' store locations** dynamically using an interactive **Leaflet map**  
- **Analyze the geographical distribution** of Starbucks stores using **choropleth heat maps**  
- Implement **dynamic filtering options** based on **country and ownership type**  
- Utilize **Shiny's reactive programming** to allow real-time updates and smooth user interactions  

## **Data and Methods**  

- **Data Cleaning & Preprocessing:**  
  - Extract relevant **columns** (e.g., `countryCode`, `ownershipTypeCode`, `latitude`, `longitude`)  
  - Remove **missing values** and format addresses for consistency  
  - Convert data into a structure suitable for **Shiny's reactive framework**  

- **Interactive Components:**  
  - **Dynamic filters**: Users can select a **specific country** or ownership type  
  - **Interactive Map**: Displays stores using **Leaflet** with **popup details**  
  - **Heatmap Visualization**: Highlights **regional density** of stores  

## **Application Implementation**  

- **Shiny UI Components:**  
  - **Dropdown filters** for selecting **country & ownership type**  
  - **Tab-based navigation** for **store location map & heatmap**  
  - **Store table output** displaying filtered results  

- **Shiny Server Logic:**  
  - **Reactive data filtering**: Updates displayed stores based on user inputs  
  - **Map interactions**: Users can zoom in/out, pan, and interact with store locations  
  - **Color-coded heatmap**: Represents store density per country  

## **Findings & Insights**  

- The **United States** has the highest number of Starbucks stores, followed by **China and Canada**.  
- There is a **higher concentration of company-owned stores (CO) in North America**, whereas **licensed stores (LS) dominate in Asia & the Middle East**.  
- The **choropleth heatmap** effectively visualizes the Starbucks store density across the world.  

## **Conclusion**  

This project successfully leverages **Shiny's interactivity** to explore and analyze **Starbucks' global presence**.  
Future improvements could include:  
- **Time-series analysis** to track **Starbucks' expansion trends**  
- **Integration with financial data** to examine store profitability by region  
- **Enhancing UI/UX** with additional **data visualizations**  


